STJ2012 analysis

### 2016-07-31 analysis/STJ2012_analysis.Rmd
add 100% OTU dataset

---
title: "STJ2012 analysis"
author: "Ross Cunning"
date: "July 28, 2016"
output: 
  html_document: 
    toc: yes
---

@@

options(scipen = 4, digits = 2)
library(phyloseq); library(vegan); library(multcompView)

@@

```{r data_import, cache=T}
# Import OTU table
OTU <- otu_table(read.table("data/otu_table.tsv", header=T, check.names=F, row.names=1,
                            skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
# Import taxonomy
TAX <- tax_table(as.matrix(read.table("data/tax_assignments.txt", sep="\t", row.names=2,
                                      col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
# Import sample data
SAM <- sample_data(read.delim("data/mapping_file.txt", sep="\t", header=T, row.names=1))
# Build phyloseq object
phy <- phyloseq(OTU, TAX, SAM)
phy
```

#### Descriptive statistics  
* Total count in OTU table: `r sum(otu_table(phy))`  
* Number of samples: `r nsamples(phy)`  
* Number of OTUs: `r ntaxa(phy)`

#### Distribution of OTU counts and reads per sample  
```{r phy_histograms, fig.height=2, fig.width=8}
taxhist <- hist(log10(taxa_sums(phy)), plot=F)
samhist <- hist(log10(sample_sums(phy)), plot=F)

par(mfrow=c(1, 2), mar=c(3,3,1,1))
plot(taxhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(samhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
```

* Range of OTU counts: `r range(taxa_sums(phy))`  
* Number of singletons:  `r length(taxa_sums(phy)[taxa_sums(phy) <= 1])`  
* Range of reads per sample: `r range(sample_sums(phy))`  
* Mean (±SD) reads per sample (arithmetic): `r mean(sample_sums(phy))` ± `r sd(sample_sums(phy))`  
* Mean (±SD) reads per sample (geometric): `r exp(mean(log(sample_sums(phy))))` ± `r exp(sd(log(sample_sums(phy))))`

```{r reads_per_sample, fig.height=2, fig.width=8}
plot_bar(phy, x="Sample", y="Abundance")
```

## Filter dataset
#### Filter samples by minimum count
```{r filter_samples}
# Set threshold number of reads
sn <- 10
# Remove samples with fewer reads than threshold
phy.f <- prune_samples(sample_sums(phy)>=sn, phy)

@@

* Number of samples remaining in dataset: `r nsamples(phy.f) `

#### Filter OTUs by minimum count
```{r filter_taxa}
# Set threshold count
n <- 3
# Identify OTUs below threshold count
taxa <- taxa_sums(phy.f)[which(taxa_sums(phy.f) >= n)]  
# Remove taxa below threshold count
phy.f <- prune_taxa(names(taxa), phy.f)  
```

* Number of OTUs remaining in dataset: `r ntaxa(phy.f)`

#### Descriptive statistics of filtered dataset
* Total count in OTU table: `r sum(otu_table(phy.f))`  
* Number of samples: `r nsamples(phy.f)`  
* Number of OTUs: `r ntaxa(phy.f)`

#### Distribution of OTU counts and reads of filtered dataset
```{r phy.f_histograms, fig.height=2, fig.width=8}
taxhist <- hist(log10(taxa_sums(phy.f)), plot=F)
samhist <- hist(log10(sample_sums(phy.f)), plot=F)

par(mfrow=c(1, 2), mar=c(3,3,1,1))
plot(taxhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(samhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
```

* Range of OTU counts: `r range(taxa_sums(phy.f))`  
* Number of singletons:  `r length(taxa_sums(phy.f)[taxa_sums(phy.f) <= 1])`  
* Range of reads per sample: `r range(sample_sums(phy.f))`  
* Mean (±SD) reads per sample (arithmetic): `r mean(sample_sums(phy.f))` ± `r sd(sample_sums(phy.f))`  
* Mean (±SD) reads per sample (geometric): `r exp(mean(log(sample_sums(phy.f))))` ± `r exp(sd(log(sample_sums(phy.f))))`

```{r reads_per_sample_filtered, fig.height=2, fig.width=8}
plot_bar(phy.f, x="Sample", y="Abundance")
```

## Transform count data
```{r transform}
# Convert to proportion (relative abundance)
phy.f.p <- transform_sample_counts(phy.f, function(x) x/sum(x))
# Apply transformation function
transform <- function(x) sqrt(x/sum(x))  # Set transformation function
phy.f.t <- transform_sample_counts(phy.f, transform)  # Transform data
```

@@

par(mfrow=c(1,2), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
composition(phy.f.p, col=taxcolors[data.frame(tax_table(phy.f.p))[order(data.frame(tax_table(phy.f.p))$Subtype), ]$Clade], legend=T)
composition(phy.f.p, col=rainbow(ntaxa(phy.f.p)), legend=F)

@@

@@

```{r strigosa}
dstr <- subset_samples(phy.f.p, Species=="strigosa")
plot_bar(dstr, x="Sample", y="Abundance", fill="OTU")
plot_bar(dstr, x="Sample", y="Abundance", fill="Subtype")
# D strigosa hosted B1, B24, B35, B38, and C3 in Finney et al.... B25, B35, and B38 are missing from ref database.
# Also try 100% OTUs
```

>>>>

---
title: "STJ2012 analysis"
author: "Ross Cunning"
date: "July 28, 2016"
output: 
  html_document: 
    self_contained: no
    toc: yes
---

@@

options(scipen = 3, digits = 2)
library(phyloseq); library(vegan); library(multcompView)

@@

```{r data_import, cache=T}
# Import OTU tables
OTU97 <- otu_table(read.table("data/97_otu_table.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
OTU100 <- otu_table(read.table("data/100_otu_table.tsv", header=T, check.names=F, row.names=1,
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
# Import taxonomy
TAX97 <- tax_table(as.matrix(read.table("data/97_tax_assignments.txt", sep="\t", row.names=2,
                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
TAX100 <- tax_table(as.matrix(read.table("data/100_tax_assignments.txt", sep="\t", row.names=2,
                                         col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
# Import sample data
SAM <- sample_data(read.delim("data/mapping_file.txt", sep="\t", header=T, row.names=1))
# Build phyloseq objects
phy97 <- phyloseq(OTU97, TAX97, SAM)
phy100 <- phyloseq(OTU100, TAX100, SAM)
```

#### Descriptive statistics and distribution of OTU counts
```{r phy_stats, fig.height=4, fig.width=8}
# Compute summary statistics
stats <- function(phy) {
  return(
    data.frame(rangeotus=paste0(range(taxa_sums(phy)), collapse=" - "),
               nsingles=length(taxa_sums(phy)[taxa_sums(phy) <= 1]),
               rangerps=paste0(range(sample_sums(phy)), collapse=" - "),
               amrps=paste(as.integer(mean(sample_sums(phy))), 
                           as.integer(sd(sample_sums(phy))), sep=" ± "),
               gmrps=paste(as.integer(exp(mean(log(sample_sums(phy))))),
                           as.integer(exp(sd(log(sample_sums(phy))))), sep=" ± "))
  )
}
stats97 <- stats(phy97)
stats100 <- stats(phy100)

# Create and plot histograms
phy <- phy97
taxhist97 <- hist(log10(taxa_sums(phy97)), plot=F)
samhist97 <- hist(log10(sample_sums(phy97)), plot=F)

taxhist100 <- hist(log10(taxa_sums(phy100)), plot=F)
samhist100 <- hist(log10(sample_sums(phy100)), plot=F)

par(mfrow=c(2, 2), mar=c(3,3,1,1))
plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
```

|                                  |    97% OTUs               |    100% OTUs                |
| -------------------------------- | ------------------------- | --------------------------- |
| Total count in OTU table         | `r sum(otu_table(phy97))` |  `r sum(otu_table(phy100))` |
| Number of OTUs                   | `r ntaxa(phy97)`          |  `r ntaxa(phy100)`          |
| Range of OTU counts              | `r stats97$rangeotus`     | `r stats100$rangeotus`      |
| Number of singleton OTUs         | `r stats97$nsingles`      | `r stats100$nsingles`       |
| Number of samples                | `r nsamples(phy97)`       |  `r nsamples(phy100)`       |
| Range of reads per sample        | `r stats97$rangerps`      | `r stats100$rangerps`       |
| Mean (±SD) reads per sample (arithmetic) | `r stats97$amrps` | `r stats100$amrps`          |
| Mean (±SD) reads per sample (geometric)  | `r stats97$gmrps` | `r stats100$gmrps`          |

## Filter dataset
#### Filter samples by minimum count
```{r filter_samples}
# Set threshold number of reads
sn <- 10
# Remove samples with fewer reads than threshold
phy97.f <- prune_samples(sample_sums(phy97)>=sn, phy97)
phy100.f <- prune_samples(sample_sums(phy100)>=sn, phy100)

@@

* Number of samples remaining in 97% OTU dataset: `r nsamples(phy97.f) `
* Number of samples remaining in 100% OTU dataset: `r nsamples(phy100.f) `

#### Filter OTUs by minimum count
```{r filter_taxa}
# Set threshold count
n <- 1
# Identify OTUs below threshold count
taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]  
taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
# Remove taxa below threshold count
phy97.f <- prune_taxa(names(taxa97), phy97.f)  
phy100.f <- prune_taxa(names(taxa100), phy100.f)  
```

* Number of OTUs remaining in 97% dataset: `r ntaxa(phy97.f)`
* Number of OTUs remaining in 97% dataset: `r ntaxa(phy100.f)`

#### Filtered data: descriptive statistics and distribution of OTUs
```{r phy.f_histograms, fig.height=4, fig.width=8}
# Compute summary statistics
stats97.f <- stats(phy97.f)
stats100.f <- stats(phy100.f)

# Create and plot histograms
taxhist97 <- hist(log10(taxa_sums(phy97.f)), plot=F)
taxhist100 <- hist(log10(taxa_sums(phy100.f)), plot=F)
samhist97 <- hist(log10(sample_sums(phy97.f)), plot=F)
samhist100 <- hist(log10(sample_sums(phy100.f)), plot=F)

par(mfrow=c(2, 2), mar=c(3,3,1,1))
plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
```

|                                  |    97% OTUs                     |    100% OTUs                |
| -------------------------------- | ------------------------------- | --------------------------- |
| Total count in OTU table         | `r sum(otu_table(phy97.f))`     | `r sum(otu_table(phy100.f))`|
| Number of OTUs                   | `r ntaxa(phy97.f)`              | `r ntaxa(phy100.f)`         |
| Range of OTU counts              | `r stats97.f$rangeotus`         | `r stats100.f$rangeotus`    |
| Number of singleton OTUs         | `r stats97.f$nsingles`          | `r stats100.f$nsingles`     |
| Number of samples                | `r nsamples(phy97.f)`           | `r nsamples(phy100.f)`      |
| Range of reads per sample        | `r stats97.f$rangerps`          | `r stats100.f$rangerps`     |
| Mean (±SD) reads per sample (arithmetic) | `r stats97.f$amrps`     | `r stats100.f$amrps`        |
| Mean (±SD) reads per sample (geometric)  | `r stats97.f$gmrps`     | `r stats100.f$gmrps`        |

## Transform count data
```{r transform}
# Convert to proportion (relative abundance)
phy97.f.p <- transform_sample_counts(phy97.f, function(x) x/sum(x))
phy100.f.p <- transform_sample_counts(phy100.f, function(x) x/sum(x))
# Apply transformation function
transform <- function(x) sqrt(x/sum(x))  # Set transformation function
phy97.f.t <- transform_sample_counts(phy97.f, transform)  # Transform data
phy100.f.t <- transform_sample_counts(phy100.f, transform)
```

@@

par(mfrow=c(1,2), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
composition(phy97.f.p, col=taxcolors[data.frame(tax_table(phy97.f.p))[order(data.frame(tax_table(phy97.f.p))$Subtype), ]$Clade], legend=T)
composition(phy97.f.p, col=rainbow(ntaxa(phy97.f.p)), legend=F)

@@

```{r, cache=T}
par(mfrow=c(1,2), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
composition(phy100.f.p, col=taxcolors[data.frame(tax_table(phy100.f.p))[order(data.frame(tax_table(phy100.f.p))$Subtype), ]$Clade], legend=T)
composition(phy100.f.p, col=rainbow(ntaxa(phy100.f.p)), legend=F)
phy.f.p <- phy97.f.p
phy.f.t <- phy97.f.t
```

@@

```{r strigosa}
dstr97 <- subset_samples(phy97.f.p, Species=="strigosa")
plot_bar(dstr97, x="Sample", y="Abundance", fill="OTU")
plot_bar(dstr97, x="Sample", y="Abundance", fill="Subtype")
# D strigosa hosted B1, B24, B35, B38, and C3 in Finney et al.... B25, B35, and B38 are missing from ref database.
# Also try 100% OTUs -- must filter more first...
#dstr100.f <- subset_samples(phy100.f, Species=="strigosa")
#dstr100.f.p <- transform_sample_counts(dstr100.f, function(x) x/sum(x))
#dstr100.f.p.f <- prune_taxa(rownames(otu_table(dstr100.f.p))[which(rowSums(otu_table(dstr100.f.p)) > 0.01)],
#                            dstr100.f.p)
#  
#plot_bar(dstr100.f.p.f, x="Sample", y="Abundance", fill="OTU")
#plot_bar(dstr100, x="Sample", y="Abundance", fill="Subtype")
#sample_names(dstr100.f)
#subset_samples(dstr100, Sample=="7")
#otu_table(dstr100.f)
#dstr100.f.p <- transform_sample_counts(dstr100.f, function(x) x/sum(x))
#colSums(otu_table(dstr100.f.p)[rowSums(otu_table(dstr100.f.p)) > 0.1, ])
